Hydras \& Co.

### 2022-05-19 doc/epsilon0-chapter.tex
book re-reading

    \section*{Conclusion}

    Let us recall three results we have proved so far.
    \begin{itemize}
    \item There exists a strictly decreasing variant which maps \texttt{Hydra} into 
    the segment $[0,\epsilon_0)$ for proving the termination of any hydra battle
    \item There exists \emph{no} such variant from \texttt{Hydra} into 
    $[0,\omega^2)$, \emph{a fortiori} into $[0,\omega)$.
    \end{itemize}

    So, a  natural question is `` Does there exist any strictly decreasing variant mapping
    type \texttt{Hydra} into some interval $[0,\mu)$ (where $\mu <\epsilon_0$) for proving the termination of all hydra battles''. The next chapter is dedicated to a formal proof that there exists no such $\mu$, even if we consider a restriction to the set of ``standard'' battles.

>>>>

    % \section*{Conclusion}

    % Let us recall three results we have proved so far.
    % \begin{itemize}
    % \item There exists a strictly decreasing variant which maps \texttt{Hydra} into 
    % the segment $[0,\epsilon_0)$ for proving the termination of any hydra battle
    % \item There exists \emph{no} such variant from \texttt{Hydra} into 
    % $[0,\omega^2)$, \emph{a fortiori} into $[0,\omega)$.
    % \end{itemize}

    % So, a  natural question is `` Does there exist any strictly decreasing variant mapping
    % type \texttt{Hydra} into some interval $[0,\mu)$ (where $\mu <\epsilon_0$) for proving the termination of all hydra battles''. The next chapter is dedicated to a formal proof that there exists no such $\mu$, even if we consider a restriction to the set of ``standard'' battles.

### 2022-05-19 doc/ks-chapter.tex
book re-reading

Our proofs are  constructive and require no axioms: they are  closed terms of the CIC, and are mainly composed on function definitions and proofs of properties of these functions. 
They  share much theoretical material with Kirby and Paris', although they do not use any knowledge about Peano arithmetic nor model  theory.  The combinatorial arguments we use and implement
come from 
 an article by J.~Ketonen and R.~Solovay~\cite{KS81}, already  cited in the work
 by L.~Kirby and J.~Paris. % on the termination of Goodstein sequences and hydra battles~\cite{KP82}.
 Section $2$ of this article: ''A hierarchy of probably recursive functions'', contains a systematic study of \emph{canonical sequences}, which are closely related to
rounds of hydra battles. 
Nevertheless, they have the same global structure as the simple proofs described in
sections~\vref{omega-case} and \vref{omega2-case}. 
We invite the reader to compare the three proofs step by step, lemma by lemma.

@@

\index{maths}{Transfinite induction}
\begin{itemize}
\item If $\alpha\not=0$, then $\canonseq{\alpha}{i}<\alpha$. Thus canonical sequences can be used for proofs by transfinite induction or function definition by transfinite recursion
\item If $\lambda$ is a limit ordinal, then $\lambda$ is the least upper bound of the set 
$\{\canonseq{\lambda}{i}\;|\,i\in\mathbb{N}_1\}$

\item If $\beta<\alpha<\epsilon_0$, then there is a ``path'' from $\alpha$ to $\beta$, \emph{i.e.} a
sequence $\alpha_0=\alpha, \alpha_1, \dots, \alpha_n=\beta$, where for every $k<n$, there exists some $i_k$ such that $\alpha_{k+1}=\canonseq{\alpha_k}{i_k}$
\item Canonical sequences correspond tightly to rounds of hydra battles: if $\alpha\not=0$,
then $\iota(\alpha)$ is transformed into $\iota(\canonseq{\alpha}{i+1})$ in one round with
the replication factor $i$ (Lemma \href{../theories/html/hydras.Hydra.O2H.html\#canonS_iota_i}{Hydra.O2H.canonS\_iota\_i}).
\item From the two previous properties, we infer that whenever $\beta<\alpha<\epsilon_0$, there exists a (free) battle from $\iota(\alpha)$ to $\iota(\beta)$.
\end{itemize}

\begin{remark}
  In~\cite{KS81}, canonical sequences are defined for any ordinal $\alpha <\epsilon_0$,
by stating that if $\alpha$ is a successor ordinal $\beta+1$,  the sequence associated with 
$\alpha$ is simply the constant sequence whose terms are equal to $\beta$.
Likewise, the canonical sequence of $0$ maps any natural number to $0$.

This convention allows us to make total the function that maps any ordinal $\alpha$ and natural number $i$ to the ordinal $\canonseq{\alpha}{i}$.

\end{remark}

First, let us recall how canonical sequences are defined in~\cite{KS81}. For efficiency's sake, we decided not to implement directly K.\&S's definitions, but to define in \gallina{} simply typed structurally recursive functions which share the abstract properties which are used in the mathematical proofs\footnote{With a small difference: the $0$-th term of the canonical sequence is not the same in our development as in~\cite{KS81}.}.

@@

\paragraph*{\gaiasign} Our library~\href{../theories/html/gaia_hydras.GCanon.html}{gaia\_hydras.GCanon} transposes the  function \texttt{canon} into \gaia's type \texttt{T1}
(please see Sect.\~ref{sect:gcanon}).
\index{gaiabridge}{Canonical sequences}
\paragraph*{Remark}
In the present state of this library, the following specializations of \texttt{canon} are still used in some proofs or lemma statements. They are planned to be deprecated.

@@

Note the use of \coq's \texttt{sig} type in the theorem's statement, which
relates the boolean function \texttt{limitb} defined on the \texttt{T1} data-type with a constructive view of the limit of a sequence: for any $\beta<\lambda$, we can compute an item of the canonical sequence of $\lambda$ which is greater than $\beta$.
We can also state directly that $\lambda$ is a (strict) least upper bound of the elements of its canonical sequence.

@@

Let us consider a kind of accessibility problem inside $\epsilon_0$: given two ordinals $\alpha$ and $\beta$, where $\beta<\alpha<\epsilon_0$, find a \emph{path} consisting of a finite sequence $\gamma_0=\alpha,\dots,\gamma_l=\beta$,
where, for every $i<l$, $\gamma_i \not= 0$ \footnote{This condition allows us to ignore paths which end by a lot of useless $0$s.} and there exists some strictly positive integer $s_i$
such that $\gamma_{i+1}=\canonseq{\gamma}{s_i}$.

@@

Note that, given $\alpha$ and $\beta$, where $\beta < \alpha$, the sequence $s$ which leads from $\alpha$ to $\beta$ is not unique.
Indeed, if $\alpha$ is a limit ordinal, the first element of $s$ can be any integer $i$ such that $\beta<\canonseq{\alpha}{i}$, and if $\alpha$ is a successor ordinal,
then the sequence $s$ can start with any positive integer.

For instance, we have also 
$\omega*2 \xrightarrow[3,4,5,6]{}\omega$. 
Likewise,
$\omega*2 \xrightarrow[1,2,1,4]{} 0$ and
$\omega*2 \xrightarrow[3,3,3,3,3,3,3,3]{} 0$.
\end{remark}

@@

The library~\href{../theories/html/gaia_hydras.GPaths.html}{gaia\_hydras.GPaths} transposes the  notion of path into \gaia's type \texttt{T1}
(please see Sect.\~ref{sect:gpath}).

@@

\noindent 
From the lemma \texttt{canon\_LT}~\vref{lemma:canon_LT}, we can convert any path into an inequality on ordinals (by induction on paths).

@@

Our goal is now  to transform any inequality $\beta<\alpha<\epsilon_0$ into a standard path $\alpha \xrightarrow[i,j]{} \beta$ for some $i$ and $j$, then into a standard battle
from $\iota(\alpha+i)$ to $\iota(\beta)$. 
Following~\cite{KS81}, we proceed in two stages:
\begin{enumerate}
\item we simulate plain (free) paths from $\alpha$ to $\beta$ with
paths made of steps $(\gamma,\canonseq{\gamma}{n})$, \emph{with the same $n$ all along the path}
\item we simulate any such path by a standard path.
\end{enumerate}

>>>>

Our proofs are  constructive and require no axioms: they are  closed terms of the CIC, and are mainly composed on function definitions and proofs of properties of these functions. 
They  borrow much theoretical material from Kirby and Paris, although they do not use any knowledge about Peano arithmetic nor model  theory.  The combinatorial arguments we use and implement
come from 
 an article by J.~Ketonen and R.~Solovay~\cite{KS81}, already  cited in the work
 by L.~Kirby and J.~Paris. % on the termination of Goodstein sequences and hydra battles~\cite{KP82}.
 Section $2$ of this article: ''A hierarchy of probably recursive functions'', contains a systematic study of \emph{canonical sequences}, which are closely related to
rounds of hydra battles. 
Nevertheless, they have the same global structure as the simple proofs described in
sections~\vref{omega-case} and \vref{omega2-case}. 
We invite the reader to compare the three proofs step by step, lemma by lemma.

@@

\index{maths}{Transfinite induction}
\begin{itemize}
\item If $\alpha\not=0$, then $\canonseq{\alpha}{i}<\alpha$. Thus canonical sequences can be used in proofs by transfinite induction or function definition by transfinite recursion
\item If $\lambda$ is a limit ordinal, then $\lambda$ is the least upper bound of the set 
$\{\canonseq{\lambda}{i}\;|\,i\in\mathbb{N}_1\}$

\item If $\beta<\alpha<\epsilon_0$, then there is a ``path'' from $\alpha$ to $\beta$, \emph{i.e.} a
sequence $\alpha_0=\alpha, \alpha_1, \dots, \alpha_n=\beta$, where for every $k<n$, there exists some $i_k$ such that $\alpha_{k+1}=\canonseq{\alpha_k}{i_k}$
\end{itemize}

\begin{remark}
  Canonical sequences are defined for any ordinal $\alpha <\epsilon_0$,
by stating that if $\alpha$ is a successor ordinal $\beta+1$,  the sequence associated with 
$\alpha$ is simply the constant sequence whose terms are equal to $\beta$.
Likewise, the canonical sequence of $0$ maps any natural number to $0$.

Thus, the function that maps any ordinal $\alpha$ and natural number $i$ to the ordinal $\canonseq{\alpha}{i}$ is total.
\end{remark}

\subsection{Canonical sequences and hydra battles}

Canonical sequences correspond tightly to rounds of hydra battles: if $\alpha\not=0$,
then $\iota(\alpha)$ is transformed into $\iota(\canonseq{\alpha}{i+1})$ in one round with
the replication factor $i$ (Lemma \href{../theories/html/hydras.Hydra.O2H.html\#canonS_iota_i}{Hydra.O2H.canonS\_iota\_i}).
Thus, whenever $\beta<\alpha<\epsilon_0$, there exists a (free) battle from $\iota(\alpha)$ to $\iota(\beta)$.

\subsection{Definitions}
First, let us recall how canonical sequences are defined in~\cite{KS81}. For efficiency's sake, we decided not to implement directly K.\&S's definitions, but to define in \gallina{} simply typed structurally recursive functions which share the abstract properties which are used in the mathematical proofs.

@@

\paragraph*{\gaiasign}
The translation of \texttt{canon} compatible with \gaia's data-types is
defined in ~\href{../theories/html/gaia_hydras.GCanon.html\#canon}{gaia\_hydras.GCanon} (please see Sect.~\ref{sect:gcanon}).

\index{gaiabridge}{Canonical sequences}
\paragraph*{Remark}
In the present state of this library, the following specializations of \texttt{canon} are still used in some proofs or lemma statements. They are planned to be deprecated.

@@

Note the use of \coq's \texttt{sig} type in the theorem's statement, which expresses a constructive view of the limit of a sequence: for any $\beta<\lambda$, we can compute an item of the canonical sequence of $\lambda$ which is greater than $\beta$.
We can also state directly that $\lambda$ is a (strict) least upper bound of the elements of its canonical sequence.

@@

Let us consider a kind of accessibility problem inside $\epsilon_0$: given two ordinals $\alpha$ and $\beta$, where $\beta<\alpha<\epsilon_0$, find a \emph{path} consisting of a finite sequence $\gamma_0=\alpha,\dots,\gamma_l=\beta$,
where, for every $i<l$, $\gamma_i \not= 0$
and there exists some strictly positive integer $s_i$
such that $\gamma_{i+1}=\canonseq{\gamma}{s_i}$.

@@

Note that, given $\alpha$ and $\beta$, where $\beta < \alpha$, the sequence $s$ which leads from $\alpha$ to $\beta$ is not unique.

For instance, we have 
$\omega\times 2 \xrightarrow[2]{}\omega$ and
$\omega\times 2 \xrightarrow[3,4,5,6]{}\omega$. 
Likewise,
$\omega\times 2 \xrightarrow[1,2,1,4]{} 0$ and
$\omega\times 2 \xrightarrow[3,3,3,3,3,3,3,3]{} 0$.
\end{remark}

@@

The library~\href{../theories/html/gaia_hydras.GPaths.html}{gaia\_hydras.GPaths} transposes the  notion of path into \gaia's type \texttt{T1}
(please see Sect.~\ref{sect:gpath}).

@@

\noindent 
By the lemma \texttt{canon\_LT}~(Sct.\vref{lemma:canon_LT}), we can convert any path into an inequality on ordinals (by induction on paths).

@@

Our goal is now  to transform any inequality $\beta<\alpha<\epsilon_0$ into a standard path $\alpha \xrightarrow[i,j]{} \beta$ for some $i$ and $j$, which corresponds to a standard battle
from $\iota(\alpha)$ (at time $i$)  to $\iota(\beta)$
(at time $j$) .
Following~\cite{KS81}, we proceed in two stages:
\begin{enumerate}
\item we simulate plain (free) paths from $\alpha$ to $\beta$ with
paths made of steps $(\gamma,\canonseq{\gamma}{n})$, \emph{with the same $n$ all along the path}.
\item we simulate any such path by a standard path.
\end{enumerate}
